A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-11 tex/codatadp.tex
special cases

\paragraph{Conventions.}
The setting is a monomorphic (or many-sorted) first-order logic.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set~$\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values $\{\bot, \top\}$. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The only predicate is equality ($\teq$) and belongs to the logical symbols.
Other predicates can be represented as functions that return $\ty{bool}$.
The metavariables $\delta,\: \tau$ range over types, with $\delta$ reserved
for datatypes and codatatypes.

@@

A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment%
\dotReportFootnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
In the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
through subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}.

@@

Datatypes and codatatypes share basic properties pertaining to constructors and
selectors. All properties below are implicitly universally quantified and range
over all $i$, $j$, $\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}
  \\
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{k-1},y,x_{k+1},\ldots,x_{n_{i\negvthinspace j}}) \longrightarrow x_k \teq y
  \\
\text{Exhaustiveness:}\quad
  & \is{C}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \is{C}_{im_i}(x)
  \\
\text{Selection:}\quad
  & s_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) = x_k
\end{aligned}
\]
%
Expressed in the algebraic jargon, exhaustiveness ensures that ``no junk''
exists, whereas distinctness and injectivity ensure that ``no confusion''
arises. The result of selectors applied to the wrong constructor is left
completely unspecified.

@@

Datatypes can be defined semantically as the initial model of an equational
specification of the selector--constructor equations. This is the view followed
by Barrett et al.\ \cite{barrett-et-al-2010}. A disadvantage with this approach
is that it does not naturally account for selectors applied to wrong
constructors. Barrett et al.\ address this by parameterizing the construction by
default values, but these give rise to spurious equalities between unrelated
terms---for example, $\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$.
This could be corrected, but the added
complexity seems to suggest that selectors are better specified axiomatically.

A related semantic view of datatypes is as initial algebra. Codatatypes are then
defined dually as a final coalgebra. The datatypes are generated by their
constructors, whereas the codatatypes are viewed through their selectors
\cite{xxx}. By uniformly focusing on the constructors, the axiomatic approach
emphasizes the commonality between datatypes and codatatypes, while sacrificing
a theoretically fruitful notion of duality.

    semantically, $k$ types are mutually (co)recursive if their dependency graph
    (x -> y if x is the type of an argument to one of y's constructors) is
    strongly connected

    (co)datatypes can be declared together as mutually (co)recursive even if
    they are not actually (co)recursive

    the semantic notion is more fruitful and the one we will focus on in the
    sequel

    Also, nothing forbids specifications with no (co)recursion, e.g.

      datatype enat = Nat(nat) | Infty
      codatatype complex = Complex(re: real, im: real)

    In this case, it makes no difference whether the type is introduced as a
    datatype or as a codatatype. Without loss of generality, we consider that
    these types are datatypes instead of codatatypes.

>>>>

\paragraph{Conventions.}
The setting is a monomorphic (or many-sorted) first-order logic.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set~$\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values $\{\bot, \top\}$. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The only predicate is equality ($\teq$) and belongs to the logical symbols.
Other predicates can be represented as functions that return $\ty{bool}$.
The metavariables $\delta,\:\varepsilon,\:\tau$ range over types, with $\delta$ reserved
for datatypes and codatatypes.

@@

\paragraph{Specifications.}
A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment%
\dotReportFootnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
In the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
through subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}.

@@

\paragraph{Properties.}
Datatypes and codatatypes share basic properties pertaining to constructors and
selectors. All properties below are implicitly universally quantified and range
over all $i$, $j$, $\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}
  \\
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{k-1},y,x_{k+1},\ldots,x_{n_{i\negvthinspace j}}) \longrightarrow x_k \teq y
  \\
\text{Exhaustiveness:}\quad
  & \is{C}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \is{C}_{im_i}(x)
  \\
\text{Selection:}\quad
  & s_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) = x_k
\end{aligned}
\]
%
Expressed in the algebraic jargon, exhaustiveness ensures that ``no junk''
exists, whereas distinctness and injectivity ensure that ``no confusion''
arises. The result of selectors applied to the wrong constructor is left
completely unspecified.

@@

\paragraph{Semantics.}
Datatypes can be defined semantically as the initial model of an equational
specification of the selector--constructor equations. This is the view followed
by Barrett et al.\ \cite{barrett-et-al-2010}. A disadvantage with this approach
is that it does not naturally account for selectors applied to wrong
constructors. Barrett et al.\ address this by parameterizing the construction by
default values, but these give rise to spurious equalities between unrelated
terms---for example, $\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$.
This could be corrected, but the added
complexity seems to suggest that selectors are better specified axiomatically.

A related semantic view of datatypes is as an initial algebra. Codatatypes are
then defined dually as a final coalgebra. The datatypes are generated by their
constructors, whereas the codatatypes are viewed through their selectors
\cite{xxx}. By uniformly focusing on the constructors, the axiomatic approach
emphasizes the commonality between datatypes and codatatypes, while
sacrificing a theoretically fruitful notion of duality.

\paragraph{Special Cases.}
A type $\delta$ depends on another type $\varepsilon$ if $\varepsilon$ is the
type of an argument to one of $\delta$'s constructors. Semantically, a set of
types is mutually (co)recursive if and only if the associated dependency graph
is strongly connected. Types can be declared together as mutually
(co)recursive even if they are not actually (co)recursive; however, the
semantic notion is more fruitful and is the one we focus on.

Furthermore, nothing forbids specifications with no (co)recursion, such as
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{enat} & \,=\, \const{Nat}(\ty{nat}) \,\mid\, \const{Infty} \\[-.5\jot]
      \keyw{codatatype}\; \,\ty{complex} & \,=\, \const{Complex}(\const{re}{:}\; \ty{real},\; \const{im}{:}\;\ty{real})
\end{aligned}
\]
Semantically, it makes no difference whether the types are introduced as
datatypes or as codatatypes.
%Without loss of generality, we consider that these types are datatypes
%instead of codatatypes.
